---
title: "Regression Models"
author: "Dan Kjeldstrøm Hansen"
date: "30 DEC 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary:
In this anaylsis I will answer the question:

Do transmission type (automatic vs. manual) have a positive effect on mpg (mile pr. gallon), and if so, how much?

The short answer to this question, and based on the following, is "yes" - given the provided dataset.
The second half of the question; "how much" is a bit more difficult to answer as several variables can cause this effect, not transmission alone. But given the dataset at hand, we will try to answer this a precise as possible.

Let's begin with a small exploration of the raw data:

## Exploratory Analysis

```{r cars}
ds <- mtcars
head(ds, 10)
str(ds)
```
The complete dataset consist of 32 different car models, with a few different variables measured for each car. (Original dataset can be found in Mosteller & Tukey (1977) along with a very indept analysis of the set and it's value as a predicting model amongs other things which is out of this small assignments idea)

For a more intuitive view, I'll factorise the am-variable:

```{r}
ds$am <- factor(ds$am,labels=c('Auto.','Man.'))
head(ds,5)
```
As we can see, transmission (am) is now presented as either Automatic or Manual
Let's see how the two type of transmission compares in regards to mpg (all other even, and given the data at hand)

```{r boxplot am}
plot(ds$am,ds$mpg,col=c('lightgreen','lightblue'),xlab='transmission', ylab='mpg')
```


## Exploratory Analysis

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```


